A full approximation storage multilevel method \\ for nonlinear variational inequalities

### 2023-06-13 paper/fascd.tex
comprehensive edits and refactor; better!

As is true for space decompositions \cite{Xu1992}, associated to a constraint decomposition (CD) are certain iterative methods \cite{Tai2003}.  Let $0\le \alpha \le 1$ be the \emph{damping parameter}.  Algorithms \ref{alg:basiccd-add} and \ref{alg:basiccd-mult} below, which we now discuss, solve smaller VI problems over each subset $\cK_i$.  These Algorithms replace $u \in \cK$ with a new iterate $w\in\cK$ which should be closer to $u^* \in \cK$ solving \eqref{eq:vi}.

@@

The additive/parallel \pr{cd-add} method (Algorithm \ref{alg:basiccd-add}) generalizes the classical Jacobi iteration.  Here ``\textbf{for all}'' can be computed in any order.  In the VI over $\mathcal{K}_i$ solved here,
\begin{equation}
\ip{f(u - \Pi_i u + \hat w_i)}{v_i-\hat w_i} \ge \ip{\ell}{v_i-\hat w_i} \qquad \forall v_i \in \mathcal{K}_i, \label{eq:cdaddvi}
\end{equation}
the expression ``$u-\Pi_iu+\hat w_i$'' removes the part of $u$ which lies in $\mathcal{K}_i$, and then adds-back the subset solution $\hat w_i \in \mathcal{K}_i$.

@@

The multiplicative/successive \pr{cd-mult} version (Algorithm \ref{alg:basiccd-mult}) generalizes the Gauss-Seidel iteration.  Each subset solution corrects the global iterate immediately, so here the ordering used in the ``\textbf{for}'' loop is important.  In the VI over $\mathcal{K}_i$,
\begin{equation}
\ip{f\Big(\sum_{j<i} w_j + \hat w_i + \sum_{j>i} \Pi_j u\Big)}{v_i-\hat w_i} \ge \ip{\ell}{v_i-\hat w_i} \qquad \forall v_i \in \mathcal{K}_i, \label{eq:cdmultvi}
\end{equation}
the sum ``$\sum_{j>i} \Pi_j u$'' should be read as ``those portions of $u$ which have not yet improved''.

@@

By \eqref{eq:constraintdecomp} and \eqref{eq:constraintrestrictionsum} one may write the test vector $v_i - \hat w_i \in \cV_i$ as a difference of admissible vectors from $\cK$.  For \eqref{eq:cdaddvi} and \eqref{eq:cdmultvi} respectively:
\begin{align}
[u - \Pi_i u + v_i] - [u - \Pi_i u + \hat w_i] &= v_i - \hat w_i, \label{eq:admissibledifferenceadd} \\
\left[\sum_{j<i} w_j + v_i + \sum_{j>i} \Pi_j u\right] - \left[\sum_{j<i} w_j + \hat w_i + \sum_{j>i} \Pi_j u\right] &= v_i - \hat w_i.  \label{eq:admissibledifferencemult}
\end{align}
In this sense the VIs over the subsets $\mathcal{K}_i$ solve the same problem as \eqref{eq:vi}.  Furthermore, let $e_i = \hat w_i - \Pi_i u \in \cV_i$.  The reader may confirm that $\hat w = u + \sum_{i} e_i$ and thus that $w = u + \alpha \sum_i e_i$.  The new iterate $w$ is computed by adding a damped correction from each sub\emph{space} $\mathcal{V}_i$, but in such a manner that $w \in \mathcal{K}$; the corrections preserve admissibility.

We call \pr{cd-add} and \pr{cd-mult} the \emph{basic CD iterations}.  VI problems \eqref{eq:cdaddvi} and \eqref{eq:cdmultvi} are analogous to constrained line searches, and in fact they reduce to such line searches when the $\cV_i$ are one-dimensional.  In general, as shown in the next Lemma \cite{Tai2003}, subject to a bound on the damping $\alpha$, which is quite restrictive for \pr{cd-add}, these iterations generate monotonically-decreasing objective values when VI problem \eqref{eq:vi} corresponds to convex optimization \eqref{eq:minimization}.

\begin{lemma}  Suppose $f=F'$ for $F:\cK\to\RR$ convex and differentiable, and let $u\in\cK$.  For $0 \le \alpha \le 1$ the output $w \in \cK$ from \pr{cd-mult} satisfies
\begin{equation}
F(w) - \ip{\ell}{w} \le F(u) - \ip{\ell}{u}.  \label{eq:objectivemonotone}
\end{equation}
If $0 \le \alpha \le 1/m$ then \pr{cd-add} likewise satisfies \eqref{eq:objectivemonotone}.
\end{lemma}

@@

To show that each iteration of the \textbf{for} loop in \pr{cd-mult} lowers the objective value, let $z_{i} = \sum_{j\le i} w_j + \sum_{j > i} \Pi_j u \in \cK$, $i=0,\dots,m-1$; it is the state at the end of the $i$th iteration.  Note that $z_{i-1} = \sum_{j < i} w_j + \Pi_i u + \sum_{j > i} \Pi_j u$ for $i>0$.  Therefore by convexity of $F$, observation \eqref{eq:admissibledifferencemult} with $v_i=\Pi_i u$, and \eqref{eq:cdmultvi} we find:
\begin{equation*}
F(z_{i-1}) - F\left(\sum_{j<i} w_j + \hat w_i + \sum_{j > i} \Pi_j u\right) \ge \ip{f\left(\sum_{j<i} w_j + \hat w_i + \sum_{j > i} \Pi_j u\right)}{\Pi_i u - \hat w_i} \ge 0.
\end{equation*}
Using convexity (again) now gives
\begin{align*}
F(z_i) &= F\left((1-\alpha) z_{i-1} + \alpha \left(\sum_{j<i} w_j + \hat w_i + \sum_{j > i} \Pi_j u\right)\right) \\
&\le  (1-\alpha) F(z_{i-1}) + \alpha F\left(\sum_{j<i} w_j + \hat w_i + \sum_{j > i} \Pi_j u\right) \le F(z_{i-1}).
\end{align*}
As $u=z_0$ and $w = z_{m-1}$, the result for \pr{cd-mult} follows.

@@

When $\alpha=1/m$ then the above proof for \pr{cd-add} can be simplified because $v=w$ and $\lambda=1$.  Then the value $F(w)$ is less than $\frac{1}{m} \sum_i F(u+e_i)$ by Jensen's inequality, which is lower than $F(u)$ by \eqref{eq:cdaddvi} or \eqref{eq:cdmultvi}, which gives inequality \eqref{eq:cdaddonereduction}.

The basic CD iterations are meaningful whether we are in the optimization case $f=F'$ or not.  Furthermore they are meaningful even when $f$ is nonlinear, non-local, or defined only on $\cK$.  However, practical and efficient FE implementation for such general operators $f$ seems not to have been addressed in the literature.  References \cite{GraeserKornhuber2009,Tai2003} only apply the CD method to the classical obstacle problem, and in that case $f=F'$ is linear, local, and defined for all inputs (i.e.~on all of $\mathcal{V}$).  The following example exposes the implementation simplifications available in such easier cases.

\begin{example}  \label{ex:fnice} Suppose $f:\cV \to \cV'$ is linear.  Furthermore suppose $f$ is a differential operator, thus local in the sense that for any $z\in\mathcal{V}$ the value $\ip{f(\phi)}{z}$ can be computed by an integral over the support of $\phi$.  Considering \pr{cd-add} for concreteness, by linearity \eqref{eq:cdaddvi} can be rewritten
\begin{equation}
\ip{f(e_i)}{v_i-\hat w_i} \ge \ip{\tilde\ell}{v_i-\hat w_i}, \label{eq:linearlocalvi}
\end{equation}
for all $v_i \in \mathcal{K}_i$, where $\tilde\ell = \ell - f(u)$.  Thus each problem \eqref{eq:linearlocalvi} can be solved using a stored residual $f(u)$ included into $\tilde\ell$.  Note $e_i = \hat w_i - \Pi_i u \in \cV_i$ is not generally admissible---generally $e_i \notin \cK$---but \eqref{eq:linearlocalvi} is meaningful because $f$ is defined over $\cV$.  If a hat function basis exists for $\cV_i$, namely a basis of small-support functions, then by locality of $f$ a solver for VI \eqref{eq:linearlocalvi} can be implemented using only integrals over these small support sets.\footnote{Compare equations (2.31) and (2.32) in \cite{Farrelletal2021}, which describe the same sense of locality.}
\end{example}

When $f$ has all the nice properties listed in Example \ref{ex:fnice}, an FE approximation of VI \eqref{eq:vi} can be solved in any incremental and efficient manner following the techniques in \cite{GraeserKornhuber2009,Tai2003}, namely by computations over hat function supports.  However, for general problems lacking the niceties of Example \ref{ex:fnice}, certain additional ideas are needed for practical implementation of \pr{cd-add} or \pr{cd-mult}.  Section \ref{sec:vcycle} constructs a multiplicative multilevel CD iteration for a general operator $f:\cK\to\cV'$ by applying the full approximation storage (FAS) idea of Brandt \cite{Brandt1977}.

@@

From now on we seek the approximate solution of certain nonlinear VI problems \eqref{eq:vi} over 2D computational domains.  Suppose $\Omega \subset \RR^2$ is a (open) bounded polygon and let $\mathcal{V}=W^{1,p}(\Omega)$.  In addition to the free boundaries generated by inequality constraints, fixed boundary conditions will be allowed, of either Dirichlet or mixed type.

We restrict to the case of inequality \emph{box constraints} \cite{BensonMunson2006,FerrisPang1997}, equivalently \emph{bi-obstacle problems}.  The \emph{lower obstacle} $\underline{\gamma}$ and the \emph{upper obstacle} $\overline{\gamma}$ are given functions, defined on the closure $\overline{\Omega}$, with values in the extended real line $\tilde\RR = [-\infty,+\infty]$.  Specifically, we suppose $\underline{\gamma} : \overline{\Omega} \to [-\infty,+\infty)$ and $\overline{\gamma} : \overline{\Omega} \to (-\infty,+\infty]$ are measurable, $\underline{\gamma} \le \overline{\gamma}$ on $\overline{\Omega}$, and $\underline{\gamma} \le g_D \le \overline{\gamma}$ on $\partial_D \Omega$.

Also, we assume that $\partial\Omega$ is split into disjoint, measurable sets, $\partial\Omega = \partial_D \Omega \cup \partial_N \Omega$, with $\partial_D \Omega$ of positive measure for simplicity to permit well-posedness \cite{Evans2010}, and $\partial_N \Omega$ possibly empty.  Suppose that the provided Dirichlet data $g_D:\partial_D \Omega \to \RR$ is sufficiently-regular so that the VI problem below is well posed.

This data defines the constraint set
\begin{equation}
\cK = \left\{w\,:\,\underline{\gamma} \le w \le \overline{\gamma} \, \text{ and }\, w\big|_{\partial_D \Omega} = g_D\right\} \subset \cV =W^{1,p}(\Omega),
\end{equation}
a closed and convex set.  (Dirichlet boundary values are in the trace sense \cite{Evans2010}.)  Note that this definition treats the inequality and equality (boundary) constraints in a unified manner.

We will solve the same VI as abstract problem \eqref{eq:vi}.  That is, given continuous $f:\cK \to \cV'$, and $\ell \in \cV'$, we assume that the following VI is well-posed for $u^*\in \cK$:
\begin{equation}
\ip{f(u^*)}{v-u^*} \ge \ip{\ell}{v-u^*} \qquad \text{for all } v\in \cK. \label{eq:boxdirichletvi}
\end{equation}
Note that $v-u^* \in W_0^{1,p}(\Omega)$, that is, test functions in \eqref{eq:boxdirichletvi} have zero values on the Dirichlet boundary $\partial_D\Omega$.  Of course, any Neumann conditions over $\partial_N \Omega$ are determined by the operator formula for $f$, which may include an integration over $\partial_N\Omega$.  In summary, VI \eqref{eq:boxdirichletvi} is both a free- and fixed-boundary value problem.

We now construct a multilevel finite element (FE) approximation to problem \eqref{eq:boxdirichletvi} using either triangular or quadrilateral meshes over $\Omega$, and $P_1$ or $Q_1$ elements \cite{Elmanetal2014}, respectively.  A hierarchy of meshes will be constructed by refinement of a coarsest mesh, with the goal being a solution on the finest-level mesh.  This hierarchy also allows the creation of a multilevel CD of $\cK$ (Section \ref{sec:cdmultilevel}).  The subsets $\cK_i$ in this CD permit coarser-level corrections in a manner which preserves admissibility while not introducing unnecessary high frequencies into the coarser-level problems, especially at the free boundary.

The mesh hierarchy and FE spaces are constructed by nested refinement in the following standard manner \cite{Elmanetal2014}.  Suppose $\mathcal{T}^0$ is the \emph{coarsest mesh}, a finite set of non-overlapping, nondegenerate elements (triangles or quadrilaterals) with union equal to $\overline{\Omega}$ and maximum edge length (characteristic mesh size) $h_0>0$.  We assume $\mathcal{T}^0$ is conforming, with no hanging nodes.  For $j=1,\dots,J$, let $\mathcal{T}^j$ be the uniform refinement of $\mathcal{T}^{j-1}$ by connecting the midpoints of edges.  For triangles and rectangles, each $\mathcal{T}^{j-1}$ element becomes four similar elements\footnote{For arbitrary triangles, and for rectangles, edge midpoint refinement generates similar elements and $h_j=2^{-j}h_0$ is the exact maximum edge length in $\mathcal{T}^j$.  For general quadrilaterals, similarity is lost and maximum edge lengths are variable; cf.~\cite{Zhang2004}.  However, only rectangular quadrilaterals are used in the Results section.} in $\mathcal{T}^j$, with characteristic mesh size $h_j = 2^{-j} h_0$.  We call $J\ge 0$ the \emph{depth} of the hierarchy, $J+1$ the \emph{number of levels}, and $\mathcal{T}^J$ the \emph{finest-level mesh}.  Let $m_j$ be the number of nodes in $\mathcal{T}^j$, so there are $m_j$ \emph{degrees of freedom} at the $j$th level.  Denote $h=h_J=2^{-J} h_0$ and $m=m_J$ for the finest level.

Let $\mathcal{V}^j$ be the continuous $P_1$ or $Q_1$ FE space \cite{Elmanetal2014} over $\mathcal{T}^j$, with arbitrary boundary values.  Observe the following vector space nesting:
\begin{equation}
\mathcal{V}^0 \subset \mathcal{V}^1 \subset \dots \subset \mathcal{V}^J \subset \mathcal{V}=W^{1,p}(\Omega).  \label{eq:fe:nestedspaces}
\end{equation}

In the VI material which follows, use of higher-order Lagrange elements $P_k$ or $Q_k$, $k\ge 2$, would imply a ``variational crime'' \cite[Chapter 10]{BrennerScott2007} of a type special to methods for inequality constraints.  Namely, our FE construction of constraint sets will require \emph{nodal monotonicity},
\begin{equation}
\bw \ge \bz \quad \implies \quad w \ge z \label{eq:nodalmonotonicity}
\end{equation}
for all $w,z \in \mathcal{V}^j$, where $\bw,\bz$ denote the vectors formed by the nodal values.  For $k\ge 2$ there are hat functions in $P_k$ and $Q_k$, i.e.~functions with nodal value one at a single node, otherwise zero, which take on negative values \cite[Figure 1.7]{Elmanetal2014}.  It is easy to show that \eqref{eq:nodalmonotonicity} is equivalent to the non-existence of indefinite-sign hat functions, so use of \eqref{eq:nodalmonotonicity} implies $k\le 1$.

Let $g_D^J \in \mathcal{V}^J$, on the finest-level mesh, satisfy $g_D^J = g_D$ on $\partial_D \Omega$.  As is typical of conforming FE methods, this restricts the Dirichlet boundary function $g_D$ to be piecewise-linear \cite{Elmanetal2014}.

To allow box constraints on all meshes in the hierarchy, let $\tilde{\mathcal{V}}^j$ denote the set of functions on the nodes of $\mathcal{T}^j$ with values in the extended real line $\tilde{\RR} = [-\infty,+\infty]$.  (Note that $\tilde{\mathcal{V}}^j$ functions are not defined everywhere on $\Omega$.)  Suppose that the finest-level \emph{lower} and \emph{upper obstacles} $\underline{\gamma}^J, \overline{\gamma}^J \in \tilde{\mathcal{V}}^J$, for nodes of $\mathcal{T}^J$, satisfy
\begin{equation}
\underline{\gamma}^J \le \overline{\gamma}^J, \qquad \underline{\gamma}^J < +\infty, \qquad -\infty < \overline{\gamma}^J. \label{eq:fe:boxconstraintrequirements}
\end{equation}
Assuming that $\underline{\gamma}^J \le g_D^J \le \overline{\gamma}^J$ on nodes of $\mathcal{T}^J$ which lie in $\partial_D \Omega$.  Then define
\begin{equation}
\mathcal{K}^J = \left\{w\,:\,\underline{\gamma}^J \le w \le \overline{\gamma}^J \text{ on nodes of } \mathcal{T}^J, \, \text{ and } \, w|_{\partial_D\Omega} = g_D|_{\partial_D\Omega}\right\} \subset \mathcal{V}^J, \label{eq:fe:fineconstraintset}
\end{equation}
as the finest-level \emph{constraint set} of admissible FE functions.  This is a nonempty, closed, and convex subset of $\mathcal{V}^J$.  As $\mathcal{V}^J$ functions never take values $\pm\infty$, inequality in \eqref{eq:fe:fineconstraintset} is strict where $\underline{\gamma}^J, \overline{\gamma}^J$ are infinite; there is no lower or upper constraint, respectively, at such nodes.

>>>>

Associated to a constraint decomposition (CD; equations \eqref{eq:subspacedecomp}--\eqref{eq:constraintrestrictionsum}) are certain iterative methods \cite{Tai2003,Xu1992}.  Let $0\le \alpha \le 1$ be the \emph{damping parameter}.  Algorithms \ref{alg:basiccd-add} and \ref{alg:basiccd-mult} below, discussed next, solve smaller VI problems over each subset $\cK_i$.  Each Algorithm replaces $u \in \cK$ with a new iterate $w\in\cK$ which should be closer to $u^* \in \cK$ solving \eqref{eq:vi}.

The additive/parallel \pr{cd-add} Algorithm \ref{alg:basiccd-add} generalizes the classical Jacobi iteration.  Here ``\textbf{for all}'' can be computed in any order.  In its VI,
\begin{equation}
\ip{f(u - \Pi_i u + \hat w_i)}{v_i-\hat w_i} \ge \ip{\ell}{v_i-\hat w_i} \qquad \forall v_i \in \mathcal{K}_i, \label{eq:cdaddvi}
\end{equation}
the expression ``$u-\Pi_iu+\hat w_i$'' removes the part of $u$ which lies in $\mathcal{K}_i$, and then adds-back the subset solution $\hat w_i \in \mathcal{K}_i$.

@@

The multiplicative/successive \pr{cd-mult} Algorithm \ref{alg:basiccd-mult} generalizes the Gauss-Seidel iteration.  Each subset solution corrects the global iterate immediately, so here the ordering used in the ``\textbf{for}'' loop is important.  In the VI,
\begin{equation}
\ip{f\Big(\sum_{j<i} w_j + \hat w_i + \sum_{j>i} \Pi_j u\Big)}{v_i-\hat w_i} \ge \ip{\ell}{v_i-\hat w_i} \qquad \forall v_i \in \mathcal{K}_i, \label{eq:cdmultvi}
\end{equation}
the sum ``$\sum_{j>i} \Pi_j u$'' should be read as ``those portions of $u$ which have not yet improved''.

@@

@@

By \eqref{eq:constraintdecomp} and \eqref{eq:constraintrestrictionsum} one may write the test vector $v_i - \hat w_i \in \cV_i$ as a difference of admissible vectors from $\cK$.  For \eqref{eq:cdaddvi} and \eqref{eq:cdmultvi} respectively:
\begin{align}
[u - \Pi_i u + v_i] - [u - \Pi_i u + \hat w_i] &= v_i - \hat w_i, \label{eq:admissibledifferenceadd} \\
\left[\sum_{j<i} w_j + v_i + \sum_{j>i} \Pi_j u\right] - \left[\sum_{j<i} w_j + \hat w_i + \sum_{j>i} \Pi_j u\right] &= v_i - \hat w_i.  \label{eq:admissibledifferencemult}
\end{align}
In this sense the VIs over the subsets $\mathcal{K}_i$ solve the same problem as \eqref{eq:vi}.  Furthermore, let $e_i = \hat w_i - \Pi_i u \in \cV_i$.  The reader may confirm that $\hat w = u + \sum_{i} e_i$ and thus that $w = u + \alpha \sum_i e_i$.  The new iterate $w$ is therefore computed by adding a damped correction from each sub\emph{space} $\mathcal{V}_i$, but in such a manner that $w \in \mathcal{K}$; the corrections preserve admissibility.

We call \pr{cd-add} and \pr{cd-mult} the \emph{basic CD iterations}.  VI problems \eqref{eq:cdaddvi} and \eqref{eq:cdmultvi} are analogous to constrained line searches, and in fact they reduce to such line searches when the $\cV_i$ are one-dimensional.  As shown in the next Lemma, when VI problem \eqref{eq:vi} corresponds to convex optimization \eqref{eq:minimization} then, subject to bounds on the damping $\alpha$, these iterations generate monotonically-decreasing objective values.

\begin{lemma} \cite{Tai2003}  Suppose $f=F'$ for $F:\cK\to\RR$ convex and differentiable, and let $u\in\cK$.  For $0 \le \alpha \le 1$ the output $w \in \cK$ from \pr{cd-mult} satisfies
\begin{equation}
F(w) - \ip{\ell}{w} \le F(u) - \ip{\ell}{u}.  \label{eq:objectivemonotone}
\end{equation}
If $0 \le \alpha \le 1/m$ then \pr{cd-add} likewise satisfies \eqref{eq:objectivemonotone}.
\end{lemma}

@@

To show that each iteration of the \textbf{for} loop in \pr{cd-mult} lowers the objective value, let $z_{i} = \sum_{j\le i} w_j + \sum_{j > i} \Pi_j u \in \cK$, $i=0,\dots,m-1$; this is the state at the end of the $i$th iteration.  Note that $z_{i-1} = \sum_{j < i} w_j + \Pi_i u + \sum_{j > i} \Pi_j u$ for $i>0$.  Therefore by convexity of $F$, observation \eqref{eq:admissibledifferencemult} with $v_i=\Pi_i u$, and \eqref{eq:cdmultvi} we find:
\begin{equation*}
F(z_{i-1}) - F\left(\sum_{j<i} w_j + \hat w_i + \sum_{j > i} \Pi_j u\right) \ge \ip{f\left(\sum_{j<i} w_j + \hat w_i + \sum_{j > i} \Pi_j u\right)}{\Pi_i u - \hat w_i} \ge 0.
\end{equation*}
Using convexity (again) now gives
\begin{align*}
F(z_i) &= F\left((1-\alpha) z_{i-1} + \alpha \left(\sum_{j<i} w_j + \hat w_i + \sum_{j > i} \Pi_j u\right)\right) \\
&\le  (1-\alpha) F(z_{i-1}) + \alpha F\left(\sum_{j<i} w_j + \hat w_i + \sum_{j > i} \Pi_j u\right) \le F(z_{i-1}).
\end{align*}
As $u=z_0$ and $w = z_{m-1}$, the result for \pr{cd-mult} follows.

@@

When $\alpha=1/m$ then the \pr{cd-add} proof can be simplified because $v=w$ and $\lambda=1$.  In this case $F(w)$ is less than $\frac{1}{m} \sum_i F(u+e_i)$ by Jensen's inequality, which is lower than $F(u)$ by \eqref{eq:cdaddvi} or \eqref{eq:cdmultvi}, which gives inequality \eqref{eq:cdaddonereduction}.

The basic CD iterations are meaningful whether or not we are in the optimization case $f=F'$, and when $f$ is nonlinear, non-local, or defined only on $\cK$.  However, practical and efficient FE implementation for such general operators $f$ seems not to have been addressed in the literature.  References \cite{GraeserKornhuber2009,Tai2003} only apply the CD method to the classical obstacle problem, and which case $f=F'$ is linear, local, and defined for all inputs (on all of $\mathcal{V}$).  Substantial implementation simplifications are available in such cases.

\begin{example}  \label{ex:fnice} Suppose $f:\cV \to \cV'$ is linear and a differential operator, thus local in the sense that for any $z\in\mathcal{V}$ the value $\ip{f(\phi)}{z}$ can be computed by an integral over the support of $\phi$.  Considering \pr{cd-add} for concreteness, by linearity \eqref{eq:cdaddvi} can be rewritten
\begin{equation}
\ip{f(e_i)}{v_i-\hat w_i} \ge \ip{\tilde\ell}{v_i-\hat w_i}, \label{eq:linearlocalvi}
\end{equation}
for all $v_i \in \mathcal{K}_i$, where $\tilde\ell = \ell - f(u)$ incorporates the initial residual $f(u)$.  Note $e_i = \hat w_i - \Pi_i u \in \cV_i$ is not generally admissible---generally $e_i \notin \cK$---but \eqref{eq:linearlocalvi} is meaningful for this $f$.
\end{example}

When $f$ has the locality of Example \ref{ex:fnice}, an FE approximation of VI \eqref{eq:vi} can be solved in any incremental and efficient manner following the techniques in \cite{GraeserKornhuber2009,Tai2003}, namely by computations over hat function supports.\footnote{Compare equations (2.31) and (2.32) in \cite{Farrelletal2021}, which describe the same sense of locality.}  However, for general problems lacking the niceties of Example \ref{ex:fnice}, certain additional ideas are needed for practical implementation of \pr{cd-add} or \pr{cd-mult}.  Section \ref{sec:vcycle} constructs a multiplicative multilevel CD iteration for a general operator $f:\cK\to\cV'$ by applying the full approximation storage (FAS) idea of Brandt \cite{Brandt1977}.

From now on we restrict to the case of \emph{box constraints} \cite{BensonMunson2006,FerrisPang1997}, equivalently \emph{bi-obstacle problems}.  Suppose $\Omega \subset \RR^d$ is open and let $\mathcal{V}=W^{1,p}(\Omega)$.  The \emph{lower obstacle} $\underline{\gamma}$ and the \emph{upper obstacle} $\overline{\gamma}$ are given functions, defined on the closure $\overline{\Omega}$, with values in the extended real line $\tilde\RR = [-\infty,+\infty]$.  Specifically, we suppose $\underline{\gamma} : \overline{\Omega} \to [-\infty,+\infty)$ and $\overline{\gamma} : \overline{\Omega} \to (-\infty,+\infty]$ are measurable, $\underline{\gamma} \le \overline{\gamma}$ on $\overline{\Omega}$, and $\underline{\gamma} \le g_D \le \overline{\gamma}$ on $\partial_D \Omega$.  Also, we assume that $\partial\Omega$ is split into disjoint, measurable sets, $\partial\Omega = \partial_D \Omega \cup \partial_N \Omega$, with $\partial_D \Omega$ of positive measure, to facilitate well-posedness \cite{Evans2010}, and $\partial_N \Omega$ possibly empty.  Dirichlet data\footnote{Dirichlet boundary values are in the trace sense \cite{Evans2010}, and $g_D$ is assumed to be as regular as needed.} $g_D:\partial_D \Omega \to \RR$ then defines
\begin{equation}
\cK = \left\{w\,:\,\underline{\gamma} \le w \le \overline{\gamma} \, \text{ and }\, w\big|_{\partial_D \Omega} = g_D\right\} \subset \cV =W^{1,p}(\Omega), \label{eq:originalconstraintset}
\end{equation}
the closed and convex constraint set.  Note that we are treating inequality and equality (boundary) constraints in a unified manner.

Recalling abstract VI problem \eqref{eq:vi}, we now consider box-constrained cases only, using set \eqref{eq:originalconstraintset}.  Given continuous $f:\cK \to \cV'$, and $\ell \in \cV'$, we will assume that the following is well-posed for $u^*\in \cK$:
\begin{equation}
\ip{f(u^*)}{v-u^*} \ge \ip{\ell}{v-u^*} \qquad \text{for all } v\in \cK. \label{eq:boxdirichletvi}
\end{equation}
Note that $v-u^* \in W_0^{1,p}(\Omega)$, so test functions in \eqref{eq:boxdirichletvi} have zero Dirichlet values.  Any Neumann conditions over $\partial_N \Omega$ are determined by the operator formula for $f$, i.e.~by an integration over $\partial_N\Omega$.  In summary, from now on we consider well-posed VI problems \eqref{eq:boxdirichletvi}, namely box-constrained boundary value problems.

@@

We seek to rapidly compute the approximate solution of box-constrained VI problems using finite elements (FE).  Suppose $\Omega \subset \RR^d$ is a (open) bounded polygon.\footnote{Section \ref{sec:results} illustrates $d=1,2,3$ cases.}  We will construct a multilevel FE approximation to problem \eqref{eq:boxdirichletvi} over $\Omega$ using $P_1$ or $Q_1$ elements \cite{Elmanetal2014}, by refining a coarsest mesh into a mesh hierarchy; the goal is an accurate solution on the finest-level mesh.  The hierarchy allows the creation of a multilevel CD of $\cK$, using the ideas in the next two Sections, which permits admissibility-preserving coarse corrections.

The mesh hierarchy and FE spaces can be constructed by standard nested refinement.\footnote{The dimension-dependent details of mesh refinement are suppressed here.  Section \ref{sec:results} only uses $P_1$ elements for intervals, triangles, and tetrahedra, in dimensions $d=1,2,3$ respectively, within the Firedrake framework \cite{Rathgeberetal2016}.}  Suppose $\mathcal{T}^0$ is the \emph{coarsest mesh}, a finite set of non-overlapping, nondegenerate elements with union equal to $\overline{\Omega}$ and maximum edge length (characteristic mesh size) $h_0>0$.  We assume $\mathcal{T}^0$ is conforming, with no hanging nodes.  For $j=1,\dots,J$, let $\mathcal{T}^j$ be the uniform refinement of $\mathcal{T}^{j-1}$ so that each $\mathcal{T}^{j-1}$ element becomes $2^d$ elements in $\mathcal{T}^j$ with maximum edge length $h_j = 2^{-j} h_0$.  We call $J\ge 0$ the \emph{depth} of the hierarchy, $J+1$ the \emph{number of levels}, and $\mathcal{T}^J$ the \emph{finest-level mesh}.  Let $m_j$ be the number of nodes in $\mathcal{T}^j$, so there are $m_j$ \emph{degrees of freedom} at the $j$th level.  Let $\mathcal{V}^j$ be the continuous $P_1$ or $Q_1$ FE space over $\mathcal{T}^j$, with arbitrary boundary values, and observe the following vector space nesting:
\begin{equation}
\mathcal{V}^0 \subset \mathcal{V}^1 \subset \dots \subset \mathcal{V}^J \subset \mathcal{V}=W^{1,p}(\Omega).  \label{eq:fe:nestedspaces}
\end{equation}
On the finest-level mesh let $g_D^J \in \mathcal{V}^J$ satisfy $g_D^J = g_D$ on $\partial_D \Omega$; this restricts $g_D$ in \eqref{eq:originalconstraintset} to be piecewise-linear.

In the VI material which follows, use of higher-order Lagrange elements, i.e.~$P_k$ or $Q_k$ with $k\ge 2$, would imply a ``variational crime'' \cite[Chapter 10]{BrennerScott2007}, special to inequality constraints, which we avoid.  Namely, our FE construction of constraint sets (Section \ref{sec:cdmultilevel}) will require \emph{nodal monotonicity},
\begin{equation}
\bw \ge \bz \quad \implies \quad w \ge z \label{eq:nodalmonotonicity}
\end{equation}
for all $w,z \in \mathcal{V}^j$, where $\bw,\bz \in \RR^{m_j}$ denote vectors of nodal values.  However, for $k\ge 2$ there are hat functions in $P_k$ and $Q_k$, i.e.~functions with nodal values one or zero, which take on negative values \cite[Figure 1.7]{Elmanetal2014}.  Evidently \eqref{eq:nodalmonotonicity} is equivalent to the non-existence of such indefinite-sign hat functions.

To allow box constraints on all meshes in the hierarchy, let $\tilde{\mathcal{V}}^j$ denote the set of functions\footnote{Note that $\tilde{\mathcal{V}}^j$ functions need not be defined everywhere on $\Omega$.} on the nodes of $\mathcal{T}^j$ with values in the extended real line $\tilde{\RR} = [-\infty,+\infty]$.  Suppose that the finest-level \emph{lower} and \emph{upper obstacles} $\underline{\gamma}^J, \overline{\gamma}^J \in \tilde{\mathcal{V}}^J$ satisfy
\begin{equation}
\underline{\gamma}^J \le \overline{\gamma}^J, \qquad \underline{\gamma}^J < +\infty, \qquad -\infty < \overline{\gamma}^J. \label{eq:fe:boxconstraintrequirements}
\end{equation}
Also assume that $\underline{\gamma}^J \le g_D^J \le \overline{\gamma}^J$ on those nodes of $\mathcal{T}^J$ which lie in $\partial_D \Omega$.  Define
\begin{equation}
\mathcal{K}^J = \left\{w\,:\,\underline{\gamma}^J \le w \le \overline{\gamma}^J \text{ on nodes of } \mathcal{T}^J, \, \text{ and } \, w|_{\partial_D\Omega} = g_D^J|_{\partial_D\Omega}\right\} \subset \mathcal{V}^J, \label{eq:fe:fineconstraintset}
\end{equation}
the finest-level \emph{constraint set} of admissible FE functions.  This is a nonempty, closed, and convex subset of $\mathcal{V}^J$.  As $\mathcal{V}^J$ functions never take values $\pm\infty$, inequality in \eqref{eq:fe:fineconstraintset} is strict where $\underline{\gamma}^J, \overline{\gamma}^J$ are infinite; there is no lower or upper constraint, respectively, at such nodes.

Let $f^J:\mathcal{K}^J \to (\mathcal{V}^J)'$ be a discretization of $f$ in \eqref{eq:boxdirichletvi}.  Note that the functional $\ell^J=\ell$ is defined and continuous over $\mathcal{V}^J$.  We seek the FE solution $u^J \in \mathcal{K}^J$ of the finest-level, finite-dimensional, nonlinear VI
\begin{equation}
\ip{f^J(u^J)}{v-u^J} \ge \ip{\ell^J}{v-u^J} \qquad \text{for all } v\in \cK^J. \label{eq:fe:vi}
\end{equation}
If $f^J$ is continuous and $p$-coercive on $\mathcal{K}^J$, in the sense that \eqref{eq:pcoercive} holds for all $u,v \in \mathcal{K}^J$, then, as for \eqref{eq:vi}, FE VI \eqref{eq:fe:vi} is well-posed.

While \eqref{eq:fe:vi} approximates \eqref{eq:boxdirichletvi}, we warn the reader that it may not technically be conforming for that VI.  That is, even if quadrature and other implementation details for $f^J$ are ignored, and despite the vector space inclusion $\mathcal{V}^J \subset \mathcal{V}$, we do not require or generally expect $\mathcal{K}^J \subset \mathcal{K}$.  Regardless of the smoothness of the continuum obstacles $\underline{\gamma}, \overline{\gamma}$, admissibility with respect to their FE interpolants $\underline{\gamma}^J, \overline{\gamma}^J$ will generally not imply continuum admissiblity.\footnote{This caveat is especially relevant in geophysical situations where only nodal values of the obstacle are known.  The nodal elevation data misses the actual topographic detail in-between the nodes.  See the Greenland ice sheet application in \cite{Bueler2016}, for example.}

The multilevel algorithm defined in the next two Sections, a solver for FE VI \eqref{eq:fe:vi}, requires transfer operators between meshes.  We require five such operators, of which three are standard multigrid tools \cite{Trottenbergetal2001}.  Table \ref{tab:transfers} will help the reader keep track.
